### 2020-09-28 text/article_vedrovice.Rmd
finish text

---
title           : "Spatial Analysis and Grave Goods Distribution at the Cemetery of Vedrovice"
# "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"

author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        :
  "Spatial patterns in cemeteries and grave goods associations at the
  cemetery of Vedrovice - Široká u Lesa are explored using computationaly
  instensive formal methods.
  Resampling approaches both to intrasite spatial arrangement and
  distribution of grave goods are explored in
  relation to layers of identity represented by sex, age and origin of the
  buried bodies.
  As a result of the investigation we conclude that...  "

@@

To determine non-randomness of various artefact combinations in burials we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original co-occurence matrix is permutated *n = 9999* times with constant
row and column totals, i.e. both the count of artifacts in individual graves
and sum of the given type of artifact across the burials are fixed.

@@

To assess relations between groups based on sex, age and origin
statistics introduced by @sosna2008 are used.

@@

Overall structure of the cemeter is examined using
<!-- either density based approaches or -->
methods based on distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
<!-- Density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence its location.-->
These methods are based on distances between pairs and triples of neighbours and
characterize the point pattern based on *second- and third-order properties*
[@nakoinz2016, 130-135; @osullivan2010, 107].
These are in contrast to the assumption underlying *first-order effects* that
any point location is independent on other point locations and
defferent parameters influence its location [@osullivan2010, 126].
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
In case of a grave placement in a burial ground we presume that the choice of
a graves location is either random or influenced predominantly by other (previous)
grave locations and
is largely independent on the outside parameters.
In other words *first-order properties* are of little to no importance in
our point of view.
The *second-order properties* and possibly *third-order properties*
are thus considered major factors in locating of the graves.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->

@@

Examining distributions of individual artefacts we find that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:cooc-vstat)).
Among the grave goods that are found to be randomly distributed in the graves
across the burial ground are ochre pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes and spondylus pendants and buckles.

@@

@@

The combinations of categories in Delaunay graph show significant relations for
pairs of bodies labeled as (1) females from maturus/senilis category with
juvenile females and (2) unsexed juveniles in general
and as already noted at several places, between (3) pairs of bodies labeled as
juveniles/non-adults.
Gabriel graph in the sex/age category shows relations between (1) adultus female and male bodies,
(2) maturus female and male bodies and (3) maturus female bodies.
The origin/sex combination of categories produces significant deviation from
expected frequencies for (1) non-local female bodies and local non-adults in
Delaunay triangulation and in Gabriel graph for (1) non-local female and local male
bodies and vice versa.
The origin/age combination shows significant departures only in Gabriel graph for
local adultus bodies with (1) non-local maturus bodies and (2) local juvenile bodies.
Last but not least, incorporating all of the categories into the networks we observe
significant results in both of the graphs (1) non-local maturus female bodies and
local juvenile (non-adult) bodies and  in case of Gabrie graph relation
between (2) non-local adultus female and local maturus male bodies.

@@

# Conclusions

@@

<!-- Artefact combinations -->

@@

Most detailed view on the clustering within the cemetery and what kinds of
bodies are prevalent in the given clusters is achieved by the percolation analysis.
The largest cluster across several percolation distances consists of more
non-adult/juvenile bodies and non-local female-sexed bodies of maturus category
on one hand and on the other the male-sexed bodies are underrepresented here.
This is in full accordance with previous observations.
The central cluster visible at small percolation distances showes overrepresentation of
local male- and female-sexed bodies.
Also the south-western cluster has higher counts of local male-sexed bodies of maturus age category and
low amount of non-local bodies.
On the other hand non-local female-sexed bodies of adultus age category are
overrepresented in the western cluster neighboring with the previously mentioned one.

@@

# Discussion

@@

<!-- randomization -->

<!--```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
# knitr::include_graphics(here("plots/s_statistic.pdf"))
```-->

@@

```{r label-cooc-nets}
label_cooc_nets <- "Networks of variables frequently present together and exluding each other, non-randomly occuring artefacts in bold"
```
```{r cooc-nets, fig.cap=label_cooc_nets }
knitr::include_graphics(here("text/figs", "cooc_nets.pdf"))
```
<!-- plans -->

```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies"}
knitr::include_graphics(here("text/figs/plan_sex.pdf"))
```

@@

```{r label-nb-sex}
label_nb_sex <- paste(
  "Mean neighbors in a given category of sex;",
  "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value"
  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
)
```
```{r nb-sex, fig.cap=label_nb_sex}
knitr::include_graphics(here("text/figs/nb_sex.pdf"))
```

```{r label-nb-age}
label_nb_age <- paste(
  "Mean neighbors in a given simplified age category"
  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
)
```
```{r nb-age, fig.cap=label_nb_age}
knitr::include_graphics(here("text/figs/nb_age.pdf"))
```

```{r label-nb-orig}
label_nb_orig <- paste(
  "Mean neighbors in a given category of origin based on Sr isotopes"
  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
)
```
```{r nb-orig, fig.cap=label_nb_orig}
knitr::include_graphics(here("text/figs/nb_orig.pdf"))
```

>>>>

---
title           : "Spatial Patterns and grave goods differences at the Cemetery of Vedrovice"
# "A resampling approach to spatial arrangement and contents of burials in the Vedrovice cemetery"

author          : "Petr Pajdla"
    # name          : "Petr Pajdla"
    # affiliation   : "Department of Archaeology and Museology,
    #                  Masaryk University, Brno, Czech Republic"
    # email         : "petr.pajdla@gmail.com"
date            : "`r format(Sys.time(), '%d %B, %Y')`"
abstract        :
  "Spatial pattern and grave goods associations at the
  cemetery of Vedrovice - Široká u Lesa are explored using computationaly
  instensive formal methods.
  Resampling approaches both to intrasite spatial arrangement and
  distribution of grave goods are explored in
  relation to layers of identity represented by different buried bodies
  attributes -- sex, age, origin and their combinations.
  We find that among non-randomly distributed grave goods
  are polished stone adzes, grinding tools, ceramic bowls and spondylus beads.
  As a result of the investigation of differences of the non-randomly distributed
  artefacts and attributes of burials we conclude that the most eminent are
  dichotomies between genders, although accentuated by both origin and age.
  Spatial analysis shows clear trends towards clustering between burials and
  the emerging clusters are further explored for similarities in burial attributes."

@@

To determine non-randomness of various artefact combinations in burials we conduct a
permutation test on the presence/absence data matrix.
The method based on generalized Monte Carlo tests was
introduced to archaeology by @manly1996 and adapted by @sosna2008.
Our approach is based on both of these works.
Original co-occurence matrix is permutated 9999 times with constant
row and column totals, i.e. both the count of artifacts in individual graves
and sum of the given type of artifact across the burials are fixed.

@@

To assess whether differences between artefacts that are non-randomly distributed
are reflected in relations between groups based on sex, age, origin and their combinations
we calculate mean Euclidean distance (ED, for details see SI) between pairs of burials in multidimensional
space settings defined by the non-randomly distributed artefacts.
The observed mean ED is compared against mean ED distribution  derived by random
reshuffling of the category labels between the burials 9999 times with constant
probabilities of appearance of the given category label.
<!-- To infer whether the observed and randomly created mean ED
distributions are samples of the same population we use non-parametric
Kruskal-Wallis test [@kruskal1952].-->
For each of the categories the deviation of the mean ED from its randomly generated
distributions is expressed as the fraction of random values of ED larger than
the one observed.
The deviation is deemed significant if less than 5% of
random values are larger than the observed value.
The process is a variation of the one described for resampling in case of
artefact combinations and similar approach is taken by @sosna2008 [346-347].

<!--
To assess relations between groups based on sex, age and origin
statistics introduced by @sosna2008 are used.-->

@@

Overall structure of the cemetery is examined using
<!-- either density based approaches or -->
methods based on distances between neighboring burials. <!--[@osullivan2010, 121-154]-->
<!-- Density based approaches, for instance quadrat tests,
characterize the point pattern in terms of its *first-order properties* [@osullivan2010, 126].
The assumption underlying the *first-order effects* is that any point location is
independent on other point locations and different parameters influence its location.-->
These methods are based on distances between pairs and triples of neighbours and
characterize the point pattern based on *second- and third-order properties*
[@nakoinz2016, 130-135; @osullivan2010, 107].
These are in contrast to the assumption underlying *first-order effects* that
any point location is independent on other point locations and
different parameters influence its location [@osullivan2010, 126].
<!-- [@nakoinz2016, 135-144; @osullivan2010, 130-139] -->
In spatial analysis of settlement patterns individual locations of
settlements (points) can depend on a number of parameters, for instance
elevation, soil type etc.
In case of a grave placement in a burial ground we presume that the choice of
a graves location is either random or influenced predominantly by other (previous)
grave locations and
is largely independent on the outside parameters.
In other words *first-order properties* are of little to no importance in
our point of view.
The *second-order properties* and possibly *third-order properties*
are thus considered major factors in locating of the graves.
<!-- It is possible that the position of burials was dependent on parameters that were not
captured by the excavation, for example location of trees but that is impossible to test.-->

@@

Examining distributions of individual artefacts we find that at the 0.05
significance level only adzes, grinding tools, ceramic bowls and spondylus beads
are distributed non-randomly.
Close to the 0.05 alpha level are distributions of bone tools and ceramic bottles
(Fig. \@ref(fig:cooc-vstat)).
Among the grave goods that are found to be randomly distributed in the graves
across the burial ground are ochre pigment, lithics, pebbles, globular pots and
special types of pottery vessel shapes and spondylus pendants and buckles.

@@

Artefacts non-randomly distributed across the cemetery are further used to
check whether significant differences exist between their occurences in
different groups defined based on origin, sex, age and their combinations.
Exploring single attributes the only differences are in the categorization to
differently sexed bodies, these are significant differences
between male- and female-sexed bodies.
In case of age or origin there are no significant differences observed.
If we combine sex and age categories the distinction between juvenile, adult and mature
female-sexed bodies and adult male-sexed bodies is evident.
Another observed dichotomy is present between adult males and juveniles (non-adults).
Moreless the same relations are present in case of combinations of sex and origin attributes.
Local male-sexed bodies have significantly different combinations of non-randomly
distributed artefacts from both local and non-local female-sexed bodies and non-adults.
If all the categories are combined the divisions are even more evident.
Local male-sexed bodies of adultus age category deviate in the artefact distributions
from adult local and non-local females, maturus/senilis non-local female-sexed bodies
and juveniles.

@@

The combinations of categories in Delaunay graph show significant relations for
pairs of bodies labeled as (1) females from maturus/senilis category with
juvenile females and (2) unsexed juveniles in general
and as already noted at several places, between (3) pairs of bodies labeled as
juveniles/non-adults.
Gabriel graph in the sex/age category shows relations between (1) adultus female and male bodies,
(2) maturus female and male bodies and (3) maturus female bodies.
The origin/sex combination of categories produces significant deviation from
expected frequencies for (1) non-local female bodies and local non-adults in
Delaunay triangulation and in Gabriel graph for (1) non-local female and local male
bodies and vice versa.
The origin/age combination shows significant departures only in Gabriel graph for
local adultus bodies with (1) non-local maturus bodies and (2) local juvenile bodies.
Last but not least, incorporating all of the categories into the networks we observe
significant results in both of the graphs (1) non-local maturus female bodies and
local juvenile (non-adult) bodies and  in case of Gabriel graph relation
between (2) non-local adultus female and local maturus male bodies.

@@

# Conclusions and discussion

@@

<!-- Artefact combinations -->
In case of grave goods analysis we explored the associations between artefacts and
different categories of the bodies not to establish or reveal any new relations
but merely to explore whether the known associations between sexes and specific grave goods
are influenced by factors such as age and local or non-local origin of the buried person.
We did not explore the relations between grave goods and burial attributes at the
individual artefact scale so to say but used an overall measure (Euclidean distance)
of between burial distance in terms of non-randomly distributed artefacts.
By establishing whether the mean ED between categories defined by the grouping attributes
mentioned above falls within or defies the expected distribution derived by
randomization experiment we identified deviations that are not random.

The most often reappearing is the difference between local adult male-sexed bodies and
both local and non-local female-sexed ones and unsexed juveniles.
This relation shows that sex classification of the body, although being
accentuated by local origin and age category, eminently structures the artefact assemblage
in the grave.

@@

Most detailed view on the clustering within the cemetery and what kinds of
bodies are prevalent in the given clusters is achieved by the percolation analysis.
The largest cluster across several percolation distances consists of more
non-adult/juvenile bodies and non-local female-sexed bodies of maturus category
on one hand and on the other the male-sexed bodies are underrepresented here.
This is in full accordance with previous observations.
The central cluster visible at small percolation distances shows overrepresentation of
local male- and female-sexed bodies.
Also the south-western cluster has higher counts of local male-sexed bodies of maturus age category and
low amount of non-local bodies.
On the other hand non-local female-sexed bodies of adultus age category are
overrepresented in the western cluster neighboring with the previously mentioned one.

@@

<!-- # Discussion -->

We have successfully shown that all of the categories considered
to be somehow related to identity in the Neolithic and especially their combinations
play substantive roles in structuring the assemblage of grave goods deposited
within the burials.
Some of the relations between the explored categories and
artefacts are well known across the LBK burial grounds,
for instance the association between
male burials and polished stone adzes [e.g. @bickle2020; @bentley2012] and some
are known specifically from Vedrovice - *Široká u lesa* cemetery, for
example the dichotomy
between globular pottery shapes being associated with female-sexed bodies and
on the other hand bowls with male-sexed bodies [@kvetina2004].
Our goal was not to establish any new connections between genders or other burial
attributes and grave good categories, but to use these associations
to see which groups of burials are well separated based on the grave goods.

First of all, we have shown that overall distribution of artefacts in the graves
was not random, the grave goods distributed non-randomly were identified and
subsequently only these were used in the further analysis effectively removing any
possible *noise* from artefacts distributed randomly.
Our main premise here is that the identity of the dead is based mostly on the attributes like sex,
age and origin defined on the basis of Strontium isotope signal.
We view these *layers* of identity as responsible for differences in
artefact compositions of different graves as well as to a certain degree a structuring
factor for the overall spatial organization of the cemetery.
In case of the grave goods this relationship is largely dominated by a category
of adult local male-sexed bodies that are distinctly different from both female-sexed bodies
of all the age categories and juveniles.
This is caused by already mentioned artefact characteristic of male burials, the stone adze.
@bickle2020 removes polished and beveled stone tools from her analysis completely
in order to remove their dominance over other patterns.
In our case, adzes are one of four non-randomly occuring artefacts and removing
them from the analysis completely would render it difficult to note any
of the visible patterns.

In case of spatial analysis we explore several approaches.
The nearest-neighbor methods are a great tool in analysing the overall cemetery but
in the small scale they do not prove useful to discover any interesting relationships
between studied *layers* of identity.
By far the most useful method in analysing the intra-site spatial arrangement
is the percolation analysis.
Another important point is the use of resampling methods that allows for identification
of random and non random patterns both in grave goods analysis and spatial organization.

<!--Traditional GIS based methods were already applied to study the cemetery
[@kvetina2004] the techniques applied here reveal some trends that were previously-->

We have shown that not only sex that is traditionally percieved
as the most important structuring proxy for identity differentiation but also
age and quite importantly also origin of the deceased played substantial
roles in equiping the graves with varius types of artefacts by the Vedrovice community.
In regard to spatial arrangement of the cemetery we demonstrated that there are
pevalent clustering procesess taking place across the cemetery and also groups
based on certain identities tend to be buried in clusters.
These observations are based on extensive use of resampling approaches that
allowed us to distinguish between patterns occuring solely by chance and
patterns significantly deviating from randomness.

@@

<!-- randomization -->
```{r sstat, fig.cap = "Distribution of simulated and observed *S* statistic"}
knitr::include_graphics(here("plots/cooc_sstat.pdf"))
```

@@

```{r label-cooc-nets}
label_cooc_nets <- "Networks of variables frequently present together and exluding each other, non-randomly occuring artefacts in bold"
```
```{r cooc-nets, fig.cap=label_cooc_nets }
knitr::include_graphics(here("text/figs", "cooc_nets.pdf"))
```

<!-- plans -->
```{r plan-sex, fig.cap="Plan of the cemetery, kernel density estimation for differently sexed bodies"}
knitr::include_graphics(here("text/figs/plan_sex.pdf"))
```

@@

<!--
```{r label-nb-sex}
label_nb_sex <- paste(
  "Mean neighbors in a given category of sex;",
  "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value"
  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
)
```
```{r nb-sex, fig.cap=label_nb_sex}
# knitr::include_graphics(here("text/figs/nb_sex.pdf"))
```

```{r label-nb-age}
label_nb_age <- paste(
  "Mean neighbors in a given simplified age category"
  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
)
```
```{r nb-age, fig.cap=label_nb_age}
# knitr::include_graphics(here("text/figs/nb_age.pdf"))
```

```{r label-nb-orig}
label_nb_orig <- paste(
  "Mean neighbors in a given category of origin based on Sr isotopes"
  # "*Gray area* -- density estimation for the simulated data, *vertical line* -- observed value;",
  # "*F* -- female-sexed body, *M* -- male-sexed body, *n. a.* -- non-adult body"
)
```
```{r nb-orig, fig.cap=label_nb_orig}
# knitr::include_graphics(here("text/figs/nb_orig.pdf"))
```
-->
